Embracing the Laws of Physics: \\ Reversible Models of Computation

### 2017-12-22 survey/survey.tex
Data II. Pi finally mostly defined. And justified. Just need to do the same for the combinators and comment on the choices. Probably add some examples too.

@@

\begin{defn}
  A \emph{commutative semiring} (sometimes called a \emph{commutative
    rig} (commutative ri\emph{n}g without negative elements) consists of a
  set $R$, two distinguished elements of $R$ named 0 and 1, and two
  binary operations~$+$ and $\cdot$, satisfying the following
  relations for any $a,b,c \in R$:
\[\begin{array}{rcl}
0 + a &=& a \\
a + b &=& b + a \\
a + (b + c) &=& (a + b) + c \\
\\
1 \cdot a &=& a \\
a \cdot b &=& b \cdot a \\
a \cdot (b \cdot c) &=& (a \cdot b) \cdot c \\
\\
0 \cdot a &=& 0 \\
(a + b) \cdot c &=& (a \cdot c) + (b \cdot c)
\end{array}\]
\end{defn}

@@

If we replace $=$ by $\simeq$, then we can indeed prove that
types (with $\bot, \top, \uplus, \times$) form a semiring.

The reader familiar with universal algebra should pause
and ponder a bit about what we have done. We have lifted 
\emph{equality} from being in the signature of the ambient logic
and instead put it in the signature of the algebraic structure
of interest. Then, and only then, are we in a solid position
to show that we have a concept (semiring) which is flexible 
enough for types to be an instance.

If we revisit the Curry-Howard correspondance, we notice one
more issue. In logic, it is true that $A \lor A = A$ and
$A \land A = A$. However, neither $A \uplus A$ nor $A \times A$ are
equivalent to $A$. They are however \emph{equi-inhabited}. This is
a fancy way of saying
\[ A \uplus A \ \text{is inhabited} \qquad \Leftrightarrow
  \qquad A \ \text{is inhabited} \]
The above is the real \textit{essence} of the correspondance.

But mere inhabitation falls far short of our goals of being
able to smoothly deform from one type to another.

Let's thus analyze the crux of the ``problem''. In logic, we have
that $\land$ and $\lor$ are both \emph{idempotent}: this is the
property of any binary operation $*$ where
$\forall a. a * a = a$. And it should be clear that idempotent
operations are \emph{forgetful}: its input has two copies of $a$,
but its output, only one. On the type side, something more subtle
happens. Consider $\top \uplus \top$ versus $\top$; the first has
\emph{two} proofs of inhabitation ($\texttt{inl tt} and \texttt{inr tt}$)
while the second only one ($\texttt{tt}$). These cannot be put in
bijective correspondance. Even though the ``payload'' \texttt{tt}
is the same, forgetting $\texttt{inl}$ (or \texttt{inr}) throws
away information -- something we have expressly disallowed.
Yes, this should remind you of Maxwell's demon: even though the
data is the same, they are tagged differently, and this
information is indeed information, and must be preserved.

@@

Introduce the usual types. Curry-Howard, as usual. Problem with it, i.e.
it is merely about inhabitation, which is NOT reversible. Show examples
with A+A and A*A. Our deformations need to be reversible, so they need
to induce type isomorphisms. So while our types are as usual, the
first level of transformations on data will change, and be isomorphisms.
The type-erasure will, naturally, be permutations. But where do the
isomorphisms come from?  Well, inhabitants of types with (0,1,+,*)
form a semiring. So, still inspired by Curry-Howard, look to the
proof-terms of semiring identities -- and voila, a nice set of
isomorphisms! We can name these isomorphisms, AND give them
implementations. Gives us a nice, reversible PL for
structured finite types. We can it Pi.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Macros

\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\idv}[3]{#2 \xrightarrow{#1} #3}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
%% \newcommand{\lid}{\textsf{lid}}
\newcommand{\alt}{~|~}
%% \newcommand{\rid}{\textsf{rid}}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\mathit{unite}_+\mathit{l}}
\newcommand{\identrp}{\mathit{uniti}_+\mathit{l}}
\newcommand{\identlsp}{\mathit{unite}_+\mathit{r}}
\newcommand{\identrsp}{\mathit{uniti}_+\mathit{r}}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{unite}_*\mathit{l}}
\newcommand{\identrt}{\mathit{uniti}_*\mathit{l}}
\newcommand{\identlst}{\mathit{unite}_*\mathit{r}}
\newcommand{\identrst}{\mathit{uniti}_*\mathit{r}}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\absorbr}{\mathit{absorbr}}
\newcommand{\absorbl}{\mathit{absorbl}}
\newcommand{\factorzr}{\mathit{factorzr}}
\newcommand{\factorzl}{\mathit{factorzl}}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\distl}{\mathit{distl}}
\newcommand{\factorl}{\mathit{factorl}}
\newcommand{\distz}{\mathit{absorbr}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

\begin{defn}
  A \emph{commutative semiring} (sometimes called a \emph{commutative
    rig} (commutative ri\emph{n}g without negative elements) consists of a
  set $R$, two distinguished elements of $R$ named 0 and 1, and two
  binary operations~$+$ and $\cdot$, satisfying the following
  relations for any $a,b,c \in R$:
\begin{equation}
\begin{array}{rcl}
0 + a &=& a \\
a + b &=& b + a \\
a + (b + c) &=& (a + b) + c \\
\\
1 \cdot a &=& a \\
a \cdot b &=& b \cdot a \\
a \cdot (b \cdot c) &=& (a \cdot b) \cdot c \\
\\
0 \cdot a &=& 0 \\
(a + b) \cdot c &=& (a \cdot c) + (b \cdot c)
\end{array}
\label{eq:csemiring}
\end{equation}
\end{defn}

@@

If we replace $=$ by $\simeq$, then we can indeed prove that
types (with $\bot, \top, \uplus, \times$) form a commutative semiring.

The reader familiar with universal algebra should pause
and ponder a bit about what we have done. We have lifted 
\emph{equality} from being in the signature of the ambient logic
and instead put it in the signature of the algebraic structure
of interest. Then, and only then, are we in a solid position
to show that we have a concept (commutative semiring) which is flexible 
enough for types to be an instance.

If we revisit the Curry-Howard correspondance, we notice one
more issue. In logic, it is true that $A \lor A = A$ and
$A \land A = A$. However, neither $A \uplus A$ nor $A \times A$ are
equivalent to $A$. They are however \emph{equi-inhabited}. This is
a fancy way of saying
\[ A \uplus A \ \text{is inhabited} \qquad \Leftrightarrow
  \qquad A \ \text{is inhabited} \]
The above is the real \textit{essence} of the correspondance.
In other words, classical Curry-Howard tells us about
\emph{logical equivalence} of types. This is even a
constructive statement: there are indeed functions
$f : A \uplus A \rightarrow A$ and $g : A \rightarrow A \uplus A$;
however, they are not inverses.

So mere inhabitation falls far short of our goals of being
able to smoothly deform from one type to another.

Let us thus analyze the crux of the ``problem''. In logic, we have
that $\land$ and $\lor$ are both \emph{idempotent}: this is the
property of any binary operation $*$ where
$\forall a. a * a = a$. And it should be clear that idempotent
operations are \emph{forgetful}: its input has two copies of $a$,
but its output, only one. On the type side, something more subtle
happens. Consider $\top \uplus \top$ versus $\top$; the first has
\emph{two} proofs of inhabitation ($\texttt{inl tt} and \texttt{inr tt}$)
while the second only one ($\texttt{tt}$). These cannot be put in
bijective correspondance. Even though the ``payload'' \texttt{tt}
is the same, forgetting $\texttt{inl}$ (or \texttt{inr}) throws
away information -- something we have expressly disallowed.
Yes, this should remind you of Maxwell's demon: even though the
data is the same, they are tagged differently, and this
data is indeed information, and must be preserved.

\begin{figure}[t]
\[
\begin{array}{rrcll}
\idc :& \tau & \iso & \tau &: \idc \\
\\
\identlp :&  0 + \tau & \iso & \tau &: \identrp \\
\swapp :&  \tau_1 + \tau_2 & \iso & \tau_2 + \tau_1 &: \swapp \\
\assoclp :&  \tau_1 + (\tau_2 + \tau_3) & \iso & (\tau_1 + \tau_2) + \tau_3 &: \assocrp \\
\\
\identlt :&  1 * \tau & \iso & \tau &: \identrt \\
\swapt :&  \tau_1 * \tau_2 & \iso & \tau_2 * \tau_1 &: \swapt \\
\assoclt :&  \tau_1 * (\tau_2 * \tau_3) & \iso & (\tau_1 * \tau_2) * \tau_3 &: \assocrt \\
\\
\distz :&~ 0 * \tau & \iso & 0 ~ &: \factorzl \\
\dist :&~ (\tau_1 + \tau_2) * \tau_3 & \iso & (\tau_1 * \tau_3) + (\tau_2 * \tau_3)~ &: \factor
\end{array}
\]
\caption{$\Pi$-terms.}
\label{pi-terms}
\end{figure}

Nevertheless, the Curry-Howard correspondance still has some
force. We know that the inhabitants of types formed with
with $\bot, \top, \uplus, \times$ are a commutative semiring. What we
want to know is, which types are equivalent? From a commutative semiring
perspective, this amounts to asking what terms are equal?
We have a set of generators for those equations, namely those
in~(\ref{eq:csemiring}). What we thus need is to create $8$ pairs
of mutually inverse functions which witness these identities.
For concreteness, we show the signatures in Figure~\ref{pi-terms}.

\begin{figure}[t]
\[
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
{\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
{}
\qquad
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \oplus c_2 : \tau_1 + \tau_3 \iso \tau_2 + \tau_4}}
{}
\]
\[
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \otimes c_2 : \tau_1 * \tau_3 \iso \tau_2 * \tau_4}}
{}
\]
\caption{$\Pi$-combinators.}
\label{pi-combinators}
\end{figure}

@@

The type-erasure will, naturally, be permutations. But where do the
isomorphisms come from?  Well, inhabitants of types with (0,1,+,*)
form a semiring. So, still inspired by Curry-Howard, look to the
proof-terms of semiring identities -- and voila, a nice set of
isomorphisms! We can name these isomorphisms, AND give them
implementations.

That gives us some ground terms.  Not enough for a PL! So then
we need to be able to 'combine' these. So we add 3 combinators for
3 different kinds of composition.
Gives us a nice, reversible PL for
structured finite types. We call it Pi.
